Statically typed Web programming

### 2012-12-08 papers/ecoop2013/repairing.tex
Remove acknowledgments
Proofreading. Just minor grammatical fixes.

\begin{abstract}
Rich Internet Applications involve more code on the client-side, dealing with DOM manipulation, event handling and
asynchronous calls to the server. Writing and maintaining large JavaScript code bases is challenging because this
language has several inadequacies (no static typing, no module system, verbose syntax, \etc{}). We investigate
further the definition of JavaScript as a Scala compiled embedded DSL using Lightweight Modular Staging. We define
safe and expressive abstractions for DOM manipulation, event handling and asynchronous programming that translate to
efficient JavaScript code using native APIs. We use both type-level information and staging to introduce high-level
abstractions with a few or no overhead on the generated JavaScript code.
\end{abstract}

@@

Writing large Web applications is known to be
challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}, mainly because of the heterogeneous
nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}:
writing distributed code requires its parts to be consistent together and leads to duplication unless you use the
same language in both sides, which, in the case of the Web, means using JavaScript to write the whole application.
However this language has several inadequacies making it hardly suitable for large code base (\eg{} no static typing,
no module system, verbose syntax, \etc). Some other in-browser execution environments give the opportunity to write
the client-side code in another language than JavaScript, \eg{} Java applets~\cite{Anuff96_JavaApplets}, Adobe
Flash~\cite{Curtis00_Flash} and Microsoft
Silverlight\footnote{\href{http://www.microsoft.com/silverlight/}{http://www.microsoft.com/silverlight/}}. However
these technologies have several drawbacks: they require an additional browser plugin to be installed (which may not
be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an
Apple smart phone), the page content can't naturally be referenced by search engines, and the content is not
structured in URLs that users can bookmark or share.

An increasing number of initiatives attempt to allow developers to write the client-side code in a language different
of JavaScript but that can be compiled to JavaScript (\eg{} GWT~\cite{Chaganti07_GWT}, Dart~\cite{Griffith11_Dart},
TypeScript\footnote{\href{http://www.typescriptlang.org/}{http://www.typescriptlang.org/}}, Roy~\cite{McKenna_Roy}).
Some of these languages can also be compiled to another runtime environment usable on server-side (\eg{} GWT, Dart,
Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}). These languages are
usually more suitable to write large applications either because they provide more constructs to build abstractions
(such as object orientation), or because they are statically typed, or because they add a bunch of useful concepts
that are missing in JavaScript (such as \emph{adhoc} polymorphism or namespaces). Moreover, their ability to compile
for both server and client sides usually allows developers to share some parts of code between server and client
sides for free. However this shared code is restricted to use exclusively language constructs: concepts provided by
external APIs can't be shared between server and client sides because the bindings with these environments are
different on the server and client sides.

@@

RIAs distinguish from classic Web pages by a higher degree of interactivity. In our example we want to let users zoom
in and out a map using their mouse wheel. To implement this feature we need to attach an event handler for the
\code{mousewheel} event on the DOM element containing the map. The following listing shows how to attach such an
event handler using the native JavaScript API:

@@

Updating the user interface usually means replacing a part of the DOM with another DOM fragment computed from data
fetched by an AJAX request. This requires to write how to compute the new DOM fragment in JavaScript. For instance
the following listing shows how to build a DOM tree representing a vertex in the mind map:

@@

Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow and
bothers modularization. Furthermore, native APIs are too much verbose, making the code hard to read.

@@

For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be
re-written with our DSL. The \code{createVertex} now returns an asynchronous value instead of taking a callback as
parameter. Then, the \code{for} expression allows us to use the vertex, when available, and to insert it on the user
interface. By making the \code{createVertex} function returning an asynchronous value instead of taking a callback as
a parameter, the code is easier to modularize into loosely coupled parts.

@@

An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. Because Scala’s
\code{for} notation is just a syntactic sugar for methods \code{foreach}, \code{map} and \code{flatMap}, we are
able to define our DSL by just defining these methods on \code{Rep[Future[A]]} values, with the following semantic:
\begin{itemize}
\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, eventually does something with the value when available~;
\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, eventually transforms the value when available~;
\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} eventually transforms the value when available~;
\end{itemize}

@@

In this section we show how we can define, as an embedded DSL, with a small effort, a template engine that is
statically typed, able to insert dynamic content in a safe way, that provides a powerful expression language,
requires no extra compilation step and that can be used on both client-side and server-side.

@@

The readability has been highly improved: nesting tags is just like nesting code blocks, HTML entities are
automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
DOM fragments definitions are written using functions so they compose just as functions compose. These benefits come
with no performance loss because the DSL generates code building DOM fragments using the native JavaScript API.

@@

Our DSL is equivalent as a template engine with Scala as the expression language. Making it usable on both server and
client sides is surprisingly as simple as defining another code generator for the DSL, producing Scala code.

@@

// Usage
x = 42;
inc();
var y = inc();
alert(y);
\end{lstlisting}
This code suffers from two problems: the variable \code{x} is global and if the user calls the \code{inc} function
without initializing first the \code{x} variable the behavior will be unexpected.

@@

The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids to explicitly thread the state
in functions using it or modifying it. In Scala, the \code{for} notation provides an elegant way to sequence monadic
computations, and thus to implicitly thread a state across computations, as shown in the following listing:

@@

Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to call a
specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
\emph{adhoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
the objects. However, modifying existing object prototypes is considered to be a bad
practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually code the dispatch logic, by registering
supported data types at the beginning of the program execution, as described in section 2.4.3
of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.

@@

Kotlin is a recent language written by JetBrains, that target both the JVM and JavaScript. It has libraries
addressing, at least partially, Web programming concerns and supports advanced programming language features such as
mixins, type inference, variance annotations for type parameters and pattern matching.

@@

\section{Acknowledgements}

We thank Tiark Rompf for his precious feedback.

>>>>

\begin{abstract}
Rich Internet Applications involve more code on the client-side, dealing with DOM manipulation, event handling and
asynchronous calls to the server. Writing and maintaining large JavaScript code bases is challenging because this
language has several inadequacies (no static typing, no module system, verbose syntax, \etc{}). We investigate
further the definition of JavaScript as a Scala compiled embedded DSL using Lightweight Modular Staging. We define
safe and expressive abstractions for DOM manipulation, event handling and asynchronous programming that translate to
efficient JavaScript code using native APIs. We use both type-level information and staging to introduce high-level
abstractions with few or no overhead on the generated JavaScript code.
\end{abstract}

@@

Writing large Web applications is known to be
challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}, mainly because of the heterogeneous
nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}:
writing distributed code requires its parts to be consistent together and leads to duplication unless one uses the
same language in both sides, which, in the case of the Web, means using JavaScript to write the whole application.
However this language has several inadequacies making it hardly suitable for large code base (\eg{} no static typing,
no module system, verbose syntax, \etc). Some other in-browser execution environments give the opportunity to write
the client-side code in another language than JavaScript, \eg{} Java applets~\cite{Anuff96_JavaApplets}, Adobe
Flash~\cite{Curtis00_Flash} and Microsoft
Silverlight\footnote{\href{http://www.microsoft.com/silverlight/}{http://www.microsoft.com/silverlight/}}. However
these technologies have several drawbacks: they require an additional browser plugin to be installed (which may not
be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an
Apple smart phone), the page content can't naturally be referenced by search engines, and the content is not
structured in URLs that users can bookmark or share.

An increasing number of initiatives attempt to allow developers to write the client-side code in a language different
than JavaScript but that can be compiled to JavaScript (\eg{} GWT~\cite{Chaganti07_GWT}, Dart~\cite{Griffith11_Dart},
TypeScript\footnote{\href{http://www.typescriptlang.org/}{http://www.typescriptlang.org/}}, Roy~\cite{McKenna_Roy}).
Some of these languages can also be compiled to another runtime environment usable on server-side (\eg{} GWT, Dart,
Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}). These languages are
usually more suitable to write large applications either because they provide more constructs to build abstractions
(such as object orientation), or because they are statically typed, or because they add some useful concepts
that are missing in JavaScript (such as \emph{adhoc} polymorphism or namespaces). Moreover, their ability to compile
for both server and client sides usually allows developers to share some parts of code between server and client
sides for free. However this shared code is restricted to use exclusively language constructs: concepts provided by
external APIs can't be shared between server and client sides because the bindings with these environments are
different on the server and client sides.

@@

RIAs distinguish themselves from classic Web pages by a higher degree of interactivity. In our example we want to let users zoom
in and out a map using their mouse wheel. To implement this feature we need to attach an event handler for the
\code{mousewheel} event on the DOM element containing the map. The following listing shows how to attach such an
event handler using the native JavaScript API:

@@

Updating the user interface usually means replacing a part of the DOM with another DOM fragment computed from data
fetched by an AJAX request. This requires writing how to compute the new DOM fragment in JavaScript. For instance
the following listing shows how to build a DOM tree representing a vertex in the mind map:

@@

Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow and
bothers modularization. Furthermore, native APIs are too verbose, making the code hard to read.

@@

For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be
re-written with our DSL. The \code{createVertex} now returns an asynchronous value instead of taking a callback as
parameter. Then, the \code{for} expression allows us to use the vertex, when available, and to insert it on the user
interface. By making the \code{createVertex} function return an asynchronous value instead of taking a callback as
a parameter, the code is easier to modularize into loosely coupled parts.

@@

An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. Because Scala’s
\code{for} notation is just syntactic sugar for methods \code{foreach}, \code{map} and \code{flatMap}, we are
able to define our DSL by just defining these methods on \code{Rep[Future[A]]} values, with the following semantic:
\begin{itemize}
\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, eventually does something with the value when available~;
\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, eventually transforms the value when available~;
\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} eventually transforms the value when available~;
\end{itemize}

@@

In this section we show how we can define, as an embedded DSL, with minimal effort, a template engine that is
statically typed, able to insert dynamic content in a safe way, that provides a powerful expression language,
requires no extra compilation step and that can be used on both client-side and server-side.

@@

The readability has been highly improved: nesting tags is just like nesting code blocks, HTML entities are
automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
DOM fragments definitions are written using functions so they compose just as functions compose. These benefits come
with no performance loss because the DSL generates code building DOM fragments by using the native JavaScript API.

@@

Our DSL is equivalent to a template engine with Scala as the expression language. Making it usable on both server and
client sides is surprisingly as simple as defining another code generator for the DSL, producing Scala code.

@@

// Usage
x = 42;
inc();
var y = inc();
alert(y);
\end{lstlisting}
This code suffers from two problems: the variable \code{x} is global and if the user calls the \code{inc} function
without first initializing the \code{x} variable the behavior will be unexpected.

@@

The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids explicitly threading the state
in functions using it or modifying it. In Scala, the \code{for} notation provides an elegant way to sequence monadic
computations, and thus to implicitly thread a state across computations, as shown in the following listing:

@@

Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to call a
specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
\emph{adhoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
the objects. However, modifying existing object prototypes is considered bad
practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by registering
supported data types at the beginning of the program execution, as described in section 2.4.3
of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.

@@

Kotlin is a recent language written by JetBrains, that targets both the JVM and JavaScript. It has libraries
addressing, at least partially, Web programming concerns and supports advanced programming language features such as
mixins, type inference, variance annotations for type parameters and pattern matching.

@@
